---
title: "Reproducible Research - Project 2 - Storm Data Analysis"
output: html_document
---
# Analysis on Weather Related Events Effects on People and the Economy based on Data from the U.S. National Oceanic and Atmosperic Adminstration's (NOAA)
by S. Ezidiegwu

# 1. Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

In this report, the NOAA Storm Database is explored to answer some basic questions about severe weather events. The results are displayed using Barplots to show the top 10 most fatal, injury incurring and economically damaging events. From our analysis, it can be inferred that the event type: Tornado, produced the most fatalities and injuries while Flooding incurred the highest economic damage in Dollar amounts.

# 2. Data Processing

## 2.1. Introduction
The data for this assignment comes in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

[Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

National Weather Service Storm Data [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

## 2.2. Assignment
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events.

### 2.2.1. Analysis Criteria
- Across the United States, which types of events (as indicated in the 𝙴𝚅𝚃𝚈𝙿𝙴 variable) are most harmful with respect to population health?
- Across the United States, which types of events have the greatest economic consequences?

*See Section 3*

## 2.3. Analysis Process

### 2.3.1. Loading the data
The data was downloaded and stored locally. Then it was loaded in R using the following code.
```{r warning=FALSE, message=FALSE}
setwd("/Users/sandraezidiegwu/Documents/Data Science/5_REPDATA/RepResearch/Storm Data Project/")
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "StormData", method = "curl")
storm.data <- read.table("StormData", header = TRUE, sep = ",")
library(dplyr)
library(ggplot2)
head(storm.data)
```

### 2.3.2. Data Extraction
Due to excessive and unneeded data, the data was cut to show the useful bit with the code below.
```{r}
cut.data <- select(storm.data, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
head(cut.data)
```

### 2.3.3. Damage Re-arrangement
Alphabetical characters used to signify magnitude
include “K” for thousands, “M” for millions, and “B” for billions. The code below reassigns those values to show numeric values. It further combines the property and crop damage to create a new vector of their total damage values.
```{r}
cut.data[cut.data$PROPDMG == 0,]$PROPDMGEXP <- "K"
cut.data[cut.data$CROPDMG == 0,]$CROPDMGEXP <- "K"
cut.data$PROPDMGEXP <- factor(cut.data$PROPDMGEXP, levels = c("K", "M", "B"))
cut.data$CROPDMGEXP <- factor(cut.data$CROPDMGEXP, levels = c("K", "M", "B"))

cut.data$PROPEXP[cut.data$PROPDMGEXP == "K"] <- 1000
cut.data$CROPEXP[cut.data$CROPDMGEXP == "K"] <- 1000
cut.data$PROPEXP[cut.data$PROPDMGEXP == "M"] <- 1e+06
cut.data$CROPEXP[cut.data$CROPDMGEXP == "M"] <- 1e+06
cut.data$PROPEXP[cut.data$PROPDMGEXP == "B"] <- 1e+09
cut.data$CROPEXP[cut.data$CROPDMGEXP == "B"] <- 1e+09

cut.data <- mutate(cut.data, PROP.DMG = PROPDMG*PROPEXP)
cut.data <- mutate(cut.data, CROP.DMG = CROPDMG*CROPEXP)
cut.data <- mutate(cut.data, TOTDMG = PROP.DMG + CROP.DMG)
```

### 2.3.4. Data Refinement
A lot of text values were rather jumbled up and required the gsub function to be gathered together for further analysis. Also, all NA values were removed from data. The code below helps refine the data into a new data frame 'ref.data'.
```{r}
ref.data <- select(cut.data, EVTYPE, FATALITIES, INJURIES, TOTDMG)

ref.data$EVTYPE <- toupper(ref.data$EVTYPE)
ref.data$EVTYPE <- factor(ref.data$EVTYPE)

ref.data$EVTYPE <- gsub("^\\s+|\\s+$", "", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("TSTMW", "THUNDERSTORM WIND", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("TSTM", "THUNDERSTORM", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("THUNDERSTORMW", "THUNDERSTORM WIND", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("THUNDERSTORMS", "THUNDERSTORM WIND", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("THUNDERSNOW$", "THUNDERSTORM WIND", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("THUNDERSNOW SHOWER", "THUNDERSTORM WIND", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("THUNDEERSTORM", "THUNDERSTORM", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("THUNDERESTORM", "THUNDERSTORM", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("THUDERSTORM", "THUNDERSTORM", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("THUNDERSTORM$", "THUNDERSTORM WIND", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("THUNERSTORM", "THUNDERSTORM", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("THUNDERSTORMW", "THUNDERSTORM WIND", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("TUNDERSTORM", "THUNDERSTORM", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("THUNDERTSORM", "THUNDERSTORM", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("^THUNDERSTORM WIND$", "THUNDERSTORM WINDS", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("^HEAT$", "EXCESSIVE HEAT", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("FLOOD/$", "FLASH FLOODS", ref.data$EVTYPE)
ref.data$EVTYPE <- gsub("^FLASH FLOOD$", "FLASH FLOODS", ref.data$EVTYPE)

ref.data[ref.data == "?"] <- NA
ref.data <- na.omit(ref.data)
```

### 2.3.5. Top Incident Totals by Event Type
Fatalities and Injuries constituted the most harmful to population health while total damage constituted most harmful to economic problems. As a result, for this anaylsis, by each incident, the total values for each of those colums were selected for evaluation. The code is shown below.
```{r}
fatal.type <- aggregate(ref.data$FATALITIES, by = list(ref.data$EVTYPE), FUN = sum)
names(fatal.type) <- c("EVENT TYPE", "TOTAL FATALITY")
top.fatal <- fatal.type[order(-fatal.type$`TOTAL FATALITY`),][1:10,]
top.fatal

injury.type <- aggregate(ref.data$INJURIES, by = list(ref.data$EVTYPE), FUN = sum)
names(injury.type) <- c("EVENT TYPE", "TOTAL INJURY")
top.injury <- injury.type[order(-injury.type$`TOTAL INJURY`),][1:10,]
top.injury

event.dmg <- aggregate(ref.data$TOTDMG, by = list(ref.data$EVTYPE), FUN = sum)
names(event.dmg) <- c("EVENT TYPE", "TOTAL DAMAGE")
top.dmg <- event.dmg[order(-event.dmg$`TOTAL DAMAGE`),][1:10,]
top.dmg
```

# 3. Results and Conclusions
```{r fig.width=7, fig.height=6}
par(mfrow = c(1,2), mar = c(12, 4, 3, 2), mgp = c(3,1,0), cex = 0.8)
barplot(top.fatal$`TOTAL FATALITY`, names.arg = top.fatal$`EVENT TYPE`, las = 3, col = 'magenta', main = "Highest Fatality by Events", ylab = "# of Fatalities")
barplot(top.injury$`TOTAL INJURY`, names.arg = top.injury$`EVENT TYPE`, las = 3, col = 'red', main = "Highest Injury by Events", ylab = "# of Injuries")
```

From the graph shown above, Tornados brought about the highest number of fatalities and injuries. It was followed by Excessive Heat and Thunderstorm Winds for fatalities and injuries respectively. 

```{r fig.width=7, fig.height=6}
par(mfrow = c(1,1), mar = c(12, 4, 3, 2), mgp = c(3,1,0), cex = 0.8)
barplot(top.dmg$`TOTAL DAMAGE`/(1e+9), names.arg = top.dmg$`EVENT TYPE`, las = 3, col = 'green', main = "Highest Economic Damage by Events", ylab = "Damage Cost ($ billions)")
```

As shown above, Floods caused the most economic damage at over $140B, followed by Hurricane/Typhoons at ~$70B.